---
title: "Model Assumptions"
author: "Maureen Nzilani"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Modelling as Count Data

Loading libraries 
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(tidyverse)
library(knitr)
library(leaflet)
library(sp)
library(sf)
library(corrplot)
library(ggplot2)
library(rlang)
library(gridExtra)
library(patchwork)
```
Loading the datasets
```{r}
data<-read_xlsx("C:\\Users\\mnzilani\\Desktop\\Working Folder\\New Data\\Counts\\Counts.xlsx")
# reading all sheets
sheet_numbers <- 1:8  # Sheets 1 to 8

# Read all the sheets (1 to 8) into a list
all_sheets <- lapply(sheet_numbers, function(sheet) {
  read_excel("C:\\Users\\mnzilani\\Desktop\\Working Folder\\New Data\\Counts\\Counts.xlsx", sheet = sheet)
})
# Name all the llags from the excel sheet
names(all_sheets) <- paste0("Sheet", sheet_numbers)

# Assign each sheet's data to a separate variable
Lag1 <- all_sheets[[1]]  # Data from Lag 1
Lag2 <- all_sheets[[2]]  # Data from Lag 2
Lag3 <- all_sheets[[3]]  # Data from Lag 3
Lag4 <- all_sheets[[4]]  # Data from lag 4
Lag5 <- all_sheets[[5]]  # Data from lag 5
Lag6 <- all_sheets[[6]]  # Data from lag 6
Lag7 <- all_sheets[[7]]  # Data from lag 7
Lag8 <- all_sheets[[8]]  # Data from lag 8

head(Lag1)
```
Checking For Missing Values in all the Lgs
```{r}
#Count for missing values in each sheet
count_missing_values <- function(data) {
  sum(is.na(data))  # Count the total number of NA values in the sheet
}

# Apply the function to each sheet
missing_values_count <- sapply(all_sheets, count_missing_values)

# Print the missing values for each sheet
missing_values_count
```
Getting summaries
```{r}

# Get the summary for Lag1
summary_Lag1 <- summary(Lag1)
summary_Lag1
```
Mapping the villages
```{r}

map <- leaflet(Lag1) %>%
  addTiles() %>%
  setView(lng = median(Lag1$lon), lat = median(Lag1$lat), zoom = 3)

map <- map %>% 
  addCircleMarkers(lng = Lag1$lon, 
                   lat = Lag1$lat,
                   radius = Lag1$Infections,
                   popup = paste("Infections:", Lag1$Infections),
                   fillOpacity = 0.7)

map
```

Checking Model Assumptions
```{r}
#1. Checking for Correlations for the numeric variables
# Select relevant columns for the correlation matrix
correlation_data <- Lag1[, c("Infections", "distance",  "tempmin", "tempmax", 
                             "mean_temp", "mean_humidity", "total_precip", "population_density", 
                             "Distance_to_Water_.km.", "forest_height", "Elevation", "LULC")]

# Calculate the correlation matrix
cor_matrix <- cor(correlation_data, use = "complete.obs", method = "spearman")

# Print the correlation matrix
#print(cor_matrix)

# Visualize the correlation matrix with correlation values inside the plot
corrplot(cor_matrix, 
         method = "color", 
         type = "full", 
         tl.col = "black", 
         tl.srt = 13, 
         addCoef.col = "black")
#Notes: Tempmax,tempmin,mean temp are heavily correlated with each other.
#Elevations is also strongly correlated with Tempmax,tempmin and mean temp
#Distance to water bodeis has a moderate correlation with Tempmax,tempmin,mean temp

```

Checking assumptions for possible count models: Poission, negative binomial, zero inflated models
```{r}
#For possion
#The response variable is a count
#Independence The observations must be independent of one another.
#Mean=Variance
#Linearity The log of the mean rate, log( λ ), must be a linear function of x.

# Calculate the mean of the infections
mean1 <- mean(Lag1$Infections, na.rm = TRUE)

print(mean1)

# Calculate the variance of the infections
variance1 <- var(Lag1$Infections, na.rm = TRUE)
print(variance1)


#Conclusion Poisson assumptions:mean=variance assumption violated mean>variance(under dispersion) {fit a possion model with log link}
#negative binomial assumption for over dispersion violated

```

Modelling as Individual cases

Loading necessary libraries
```{r}
library(dplyr)
library(tidyr)
library(INLA)
library(ggplot2)
library(Metrics)
library(sp)
library(spdep)
```

Fitting a Logistic regression and checking if the residuals have spatial autocorrelation
```{r}
dr<-read.csv("C:\\Users\\mnzilani\\Desktop\\Working Folder\\New Data\\Lags\\L3.csv")
dr <- dr %>%
  mutate(age_group = case_when(
    AGE >= 0 & AGE <= 4 ~ "0-4 years",
    AGE >= 4.01 & AGE <= 19 ~ "5-19 years",
    AGE >= 20 & AGE <= 60 ~ "20-60 years",
    AGE > 60 ~ "60 and above years"
  ))
modellogistic <- glm(Infections ~ SEX+Distance_to_Water_.km. + population_density + forest_height +
                       tempmin + mean_humidity + total_precip  + age_group+ distance+Distance_to_Water_.km., 
                     data = dr, family = binomial)
summary(modellogistic)
summary(modellogistic$residuals)
hist(modellogistic$residuals, main = "Histogram of Residuals", xlab = "Residuals")
# Add jitter to lat and lon 
dr <- dr %>%
  mutate(
    lat_jittered = lat + runif(n(), -1e-5, 1e-5),
    lon_jittered = lon + runif(n(), -1e-5, 1e-5)
  )

coordinates(dr) <- ~lon_jittered + lat_jittered   
# apply k-nearest neighbors 
neighbors <- knn2nb(knearneigh(coordinates(dr), k=4))
listw <- nb2listw(neighbors, style="W")
# Calculate Moran's I for the residuals
residuals <- residuals(modellogistic, type = "response")
moran_test <- moran.test(residuals, listw)

# Print the Moran's I test results
print(moran_test)
```

Checking for Spatial autocorrelation in Infections
```{r}
dr1<-read.csv("C:\\Users\\mnzilani\\Desktop\\Working Folder\\New Data\\Lags\\L3.csv")
dr1 <- dr1 %>%
  mutate(
    lat_jittered = lat + runif(n(), -1e-5, 1e-5),
    lon_jittered = lon + runif(n(), -1e-5, 1e-5)
  )

# Define coordinates 
coords <- cbind(dr1$lat_jittered, dr1$lon_jittered)
# Apply the knearneigh 
nb <- knn2nb(knearneigh(coords, k = 4))
listw <- nb2listw(nb, style = "W")

# Calculate Moran's I for spatial autocorrelation
moran_test <- moran.test(dr1$Infections,listw)

# Display the results
moran_test
```
## there is spatial autocorrelation in the residuals of the model and in infections
